# Choosing Between Alternative Models {#ch:altmods}

## Non-linearity illustrations


```{r setup-ch6, include=FALSE}
knitr::opts_knit$set(
  echo = TRUE,
  root.dir = here::here()
)

knitr::opts_chunk$set(
  fig.path = "fig/"
)

library(rms)
library(foreign)
library(lspline)
library(lattice)
library(latticeExtra)

mycolors <- c(
  "black" = 1,
  "Red" = "#ED0000",
  "CongressBlue" = "#00468B",
  "Apple" = "#42B540",
  "BondiBlue" = "#0099B4",
  "TrendyPink" = "#925E9F",
  "Carmine" = "#AD002A",
  "CodGray" = "#1B1919",
  "MonaLisa" = "#FDAF91",
  "Edward" = "#ADB6B6"
)
```

### Prepare GUSTO data
Some logistic regression fits with linear, square, rcs, linear spline terms

```{r}
# Import gusto; publicly available
gusto <- read.csv("data/gusto_age.csv")[-1]
Fmort <- as.data.frame(read.csv("data/Fmort.csv"))[-1]
Fmort$age10 	<- Fmort$age/10
Fmort$age102 	<- Fmort$age10^2
```


```{r echo=F}
dd <- datadist(gusto)
options(datadist="dd")

## Age linear; add square; rcs
agegusto.linear <- lrm(DAY30~AGE,	data=gusto, x=T,y=T)
agegusto.square <- lrm(DAY30~pol(AGE,2),	data=gusto, x=T,y=T)
agegusto.rcs    <- lrm(DAY30~rcs(AGE,5),	data=gusto, x=T,y=T)

## linear spline, age 50 (equals value 5.0 for decades)
library(lspline)
agegusto.linearspline    <- lrm(DAY30~lspline(AGE,5.0),	data=gusto, x=T,y=T)

## dichotomize
agegusto.cat65    <- lrm(DAY30~ifelse(AGE<6.5,0,1),	data=gusto, x=T,y=T)
```

### anova results for the different fits
We note minor differences between the continuous fits, and a clear loss of information for the dichtomization at age 65 years
```{r echo=T}
anova(agegusto.linear)
anova(agegusto.square)
anova(agegusto.rcs)
anova(agegusto.linearspline)
anova(agegusto.cat65)
```

### Plotting of age effects
Plot age effect first at lp scale (logodds), then at probability scale

Age effect at logodds scale;  Age effect at probability scale
```{r echo=F}
## Prepare plots, age 20 to 95 years
newdata.age =data.frame("AGE"=seq(2,9.5, by=0.01))
pred.agegusto.linear  <- predict(agegusto.linear, newdata.age)
pred.agegusto.square  <- predict(agegusto.square,newdata.age)
pred.agegusto.rcs     <- predict(agegusto.rcs,newdata.age)
pred.agegusto.linearspline <- predict(agegusto.linearspline,newdata.age)
pred.agegusto.cat65   <- predict(agegusto.cat65,newdata.age)

# Start plotting
par(mfrow=c(1,2))
plot(x=10*newdata.age[,1], y=pred.agegusto.linear, xlim=c(20,92), ylim=c(-6,0), las=1, xaxt="n",
     xlab='Age in years', ylab='logit of 30-day mortality', cex.lab=1.2, type="l", lwd=2)
axis(1, at=c(30, 50, 65, 80))
lines(x=10*newdata.age[,1], y=pred.agegusto.square, lty=2, lwd=2, col=mycolors[2])
lines(x=10*newdata.age[,1], y=pred.agegusto.rcs, lty=3, lwd=2, col=mycolors[3])
lines(x=10*newdata.age[,1], y=pred.agegusto.linearspline, lty=4,lwd=2, col=mycolors[4])
lines(x=10*newdata.age[,1], y=pred.agegusto.cat65, lty=5,lwd=2, col=mycolors[5])

legend("topleft", legend=c("linear","square","rcs","lspline","dichotomize"),
       lty=1:5, lwd=2, cex=1, bty='n',col=mycolors[1:5])

scat1d(x=10*gusto$AGE, side=1, frac=.03, col="darkblue")
title("Logit scale")

## Probability scale
newdata.age =data.frame("AGE"=seq(2,9.5, by=0.01))
pred.agegusto.linear  <- plogis(predict(agegusto.linear, newdata.age))
pred.agegusto.square  <- plogis(predict(agegusto.square,newdata.age))
pred.agegusto.rcs     <- plogis(predict(agegusto.rcs,newdata.age))
pred.agegusto.linearspline <- plogis(predict(agegusto.linearspline,newdata.age))
pred.agegusto.cat65   <- plogis(predict(agegusto.cat65,newdata.age))

plot(x=10*newdata.age[,1], y=pred.agegusto.linear, xlim=c(20,92), ylim=c(0,.5), las=1, xaxt="n",
     xlab='Age in years', ylab='probability of 30-day mortality', cex.lab=1.2, type="l", lwd=2)
axis(1, at=c(30, 50, 65, 80))
lines(x=10*newdata.age[,1], y=pred.agegusto.square, lty=2, lwd=2, col=mycolors[2])
lines(x=10*newdata.age[,1], y=pred.agegusto.rcs, lty=3, lwd=2, col=mycolors[3])
lines(x=10*newdata.age[,1], y=pred.agegusto.linearspline, lty=4, lwd=2, col=mycolors[4])
lines(x=10*newdata.age[,1], y=pred.agegusto.cat65, lty=5, lwd=2, col=mycolors[5])
scat1d(x=10*gusto$AGE, side=1, frac=.03, col="darkblue")
title("Probability scale")
```


Fig 6.1

### Start surgical mortality by age in Medicare

Age effect at logodds scale
```{r Medicare_data, echo=F, warning=F}
dd <- datadist(Fmort)
options(datadist="dd")

# Age continuous
fitage		<- lrm(mort ~ surgery + age10, weights=w, maxit=500, eps=.025, data=Fmort)
fitagei	<- update(fitage,  ~ . + surgery * age10)
fitage2	<- update(fitage,  ~ . + age102)
fitagei2	<- update(fitage,  ~ . + age102 + surgery * age10)
fitplot	<- lrm(mort ~ surgery * age, weights=w, data=Fmort, x=T, y=T)
fitplot2	<- lrm(mort ~ surgery + age, weights=w, data=Fmort, x=T, y=T)
```

### anova results for the fit of age, with interaction by type of surgery

Type of surgery is clearly most relevant (chi2 >13500) in all fits. Age is als relevant (chi2>3000), and a square term is not needed (chi2 = 2); the interaction adds a little bit (chi2 95). With these large numbers (1.1M patients), most effects have p<.0001.

We will evaluate the differences between fits with or without interaction term graphically further down

```{r echo=T}
# Look for model improvements 
anova(fitplot2)  # linear age effect, no interaction with surgery
anova(fitage2)   # age square added
anova(fitplot)   # interaction added to linear age effect
```

### Plotting of predicted age effects, with interaction by type of surgery; add 95% CI

Plot age effects at logodds scale with 95% CI

```{r echo=F}
plot(Predict(fitplot, age=65:100, surgery=NA), ylab="Mortality", 
     par.settings = list(superpose.line = list(col = mycolors[c(2:10, 2:6)])))
```

### Plotting of age effects with original data points

Fit with interaction (solid lines) and no interaction (dashed lines)

```{r echo=F}
# Points
Fsum <- aggregate(data=Fmort, w ~ age + surgery, sum) #total N
Fdead <- aggregate(data=Fmort, w ~ age + surgery, min) # N dead
Fsum$mort <- Fdead[,3] / Fsum[,3] # mortality per group
Fsum$dead <- Fdead[,3]
Fsum$lmort <- log(Fsum$mort / (1-Fsum$mort))

# Interaction fit
fit.mat <- Predict(fitplot, age=65:100, surgery=levels(Fmort$surgery))
# Main effect fit
fit.mat2 <- Predict(fitplot2, age=65:100, surgery=levels(Fmort$surgery))

# Do some advanced lattice plotting
foo_key <- list(corner=c(1,1), x=80, y=-1,
  text=list(lab=c("Observed","Age interaction", "Common age effect")),
                     points=list(pch=c(16,NA, NA), cex=c(2, NA, NA), col=c("red", "red", "black")), 
                     lines=list(type = c("p", "l", "l"), pch=16, lty=c(0,1,2), lwd=3, col=c("red", "red", "black")))
                

plot1<-xyplot(lmort ~ age | surgery, groups = surgery, Fsum,ylim=c(-4.7,-0.8),
              ylab="logit(mortality)", xlab="Age (years)", xlim=c(64,92), 
              par.settings = list(superpose.symbol = list(pch = 16, cex = 1, col = mycolors[c(2:10, 2:6)]),
                                  par.ylab.text = list(cex=1.2), par.xlab.text = list(cex=1.2)))

plot2<-xyplot(yhat ~ age | surgery, groups = surgery, fit.mat, type='l',
              ylab="logit(mortality)", xlab="Age (years)", xlim=c(64,92), 
              par.settings = list(superpose.line = list(lwd=3, col = mycolors[c(2:10, 2:6)]),
                                  par.ylab.text = list(cex=1.2), par.xlab.text = list(cex=1.2)))

plot3<-xyplot(yhat ~ age | surgery, groups = surgery, fit.mat2, type='l',
              ylab="logit(mortality)", xlab="Age (years)", xlim=c(64,92), 
              par.settings = list(superpose.line = list(lwd=2, lty=2, col = 'black'),
                                  par.ylab.text = list(cex=1.2), par.xlab.text = list(cex=1.2)))

# all three overlayed
plot1+plot2+plot3
```

