# Coding of Categorical and Continuous Predictors {#ch:coding}

```{r setup-ch9, include=FALSE}
knitr::opts_knit$set(
  echo = TRUE,
  root.dir = here::here()
)

knitr::opts_chunk$set(
  fig.path = "fig/"
)

library(rms)
library(foreign)
library(mfp)
library(gam)
options(width = 100)
options(digits = 3)

mycolors <- c(
  "black" = 1,
  "Red" = "#ED0000",
  "CongressBlue" = "#00468B",
  "Apple" = "#42B540",
  "BondiBlue" = "#0099B4",
  "TrendyPink" = "#925E9F",
  "Carmine" = "#AD002A",
  "CodGray" = "#1B1919",
  "MonaLisa" = "#FDAF91",
  "Edward" = "#ADB6B6"
)
```

## Figues 9.1 to 9.6

GUSTO-I is a data set with patients suffering from an acute myocardial infarction, where we want to predict 30-day mortality.
TBI is a data set with patients suffering from a moderate or severe traumatic brain injury.

```{r import_data, warnings = FALSE }
# Import gusto
gusto <- read.csv("data/gusto_age_STE.csv")[, -1]

# Import sample4; n=785
gustos <- read.csv("data/Gustos4Age.csv")

# Import TBI data; n=2159
TBI <- read.csv("data/TBI2vars.csv")
```

### Fig 9.1: Age linear; add square; rcs in GUSTO-I

```{r Fig9.1, message = FALSE }
agegusto.linear <- lrm(DAY30 ~ AGE, data = gusto, x = T, y = T)
agegusto.square <- lrm(DAY30 ~ pol(AGE, 2), data = gusto, x = T, y = T)
agegusto.rcs <- lrm(DAY30 ~ rcs(AGE, 5), data = gusto, x = T, y = T)
## dichotomize
agegusto.cat65 <- lrm(DAY30 ~ ifelse(AGE < 65, 0, 1), data = gusto, x = T, y = T)
## 3 categories
agegusto.3cat <- lrm(DAY30 ~ ifelse(AGE < 60, 0, ifelse(AGE < 70, 1, 2)), data = gusto, x = T, y = T)

# Predict for age 20:95
newdata.age <- data.frame("AGE" = seq(20, 95, by = 0.1))
pred.agegusto.linear <- predict(agegusto.linear, newdata.age)
pred.agegusto.square <- predict(agegusto.square, newdata.age)
pred.agegusto.rcs <- predict(agegusto.rcs, newdata.age)
pred.agegusto.cat65 <- predict(agegusto.cat65, newdata.age)
pred.agegusto.3cat <- predict(agegusto.3cat, newdata.age)

# Make plot
dd <- datadist(gusto)
options(datadist = "dd") # for rms
par(mfrow = c(1, 1))
plot(
  x = newdata.age[, 1], y = pred.agegusto.linear, xlim = c(20, 92), ylim = c(-7, 0), las = 1, xaxt = "n",
  xlab = "Age in years", ylab = "logit of 30-day mortality", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[2]
)
axis(1, at = c(40, 50, 60, 65, 70, 80, 90))
lines(x = newdata.age[, 1], y = pred.agegusto.cat65, lty = 2, lwd = 2, col = mycolors[3])
lines(x = newdata.age[, 1], y = pred.agegusto.3cat, lty = 3, lwd = 2, col = mycolors[4])
scat1d(x = gusto$AGE, side = 1, frac = .05, col = "darkblue")
legend("topleft",
  legend = c("linear", "<65 vs >=65", "<60, 60-69, 70+"),
  lty = c(1, 2, 3), lwd = 2, cex = 1, bty = "n", col = mycolors[2:4]
)
```


### Fig 9.2: Impact of number of ST elevations (STE)

```{r Fig9.2, message = FALSE }
# winsorize at STE 10
gusto$STE <- ifelse(gusto$STE > 10, 10, gusto$STE)
gusto$STE.f <- as.factor(gusto$STE)
STE.linear <- lrm(DAY30 ~ STE, data = gusto, x = T, y = T)
STE.square <- lrm(DAY30 ~ pol(STE, 2), data = gusto, x = T, y = T)
STE.rcs <- lrm(DAY30 ~ rcs(STE, 5), data = gusto, x = T, y = T)
STE.factor <- lrm(DAY30 ~ STE.f, data = gusto, x = T, y = T)

# dichotomize
STE.cat4 <- lrm(DAY30 ~ ifelse(STE < 5, 0, 1), data = gusto, x = T, y = T)

# predict for STE 0:10
newdata.STE <- data.frame("STE" = seq(0, 10, by = 1))
pred.STE.linear <- predict(STE.linear, newdata.STE)
pred.STE.square <- predict(STE.square, newdata.STE)
pred.STE.rcs <- predict(STE.rcs, newdata.STE)
pred.STE.cat4 <- predict(STE.cat4, newdata.STE)

par(mfrow = c(1, 1))
plot(
  x = newdata.STE[, 1], y = pred.STE.linear, las = 1,
  xlab = "Number of leads with ST elevation", ylab = "logit of 30-day mortality", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[2]
)
lines(x = newdata.STE[, 1], y = pred.STE.square, lty = 2, lwd = 2, col = mycolors[3])
lines(x = newdata.STE[, 1], y = pred.STE.rcs, lty = 3, lwd = 3, col = mycolors[4])
lines(x = newdata.STE[1:5, 1], y = pred.STE.cat4[1:5], lty = 4, lwd = 2, col = mycolors[5])
lines(x = newdata.STE[6:11, 1], y = pred.STE.cat4[6:11], lty = 4, lwd = 2, col = mycolors[5])

# Original data points, with size proportional to sqrt(events)
STEmort <- log(by(gusto$DAY, gusto$STE, mean) / (1 - by(gusto$DAY, gusto$STE, mean)))[1:11]
STEw <- sqrt(by(gusto$DAY, gusto$STE, sum)[1:11]) / 10
points(x = newdata.STE[, 1], y = STEmort, pch = 1, lwd = 2, cex = STEw, col = "black")
scat1d(x = gusto$STE, side = 1, frac = .05, col = "darkblue")
legend("topleft",
  legend = c("data", "linear", ".. +square", "rcs", "<=4, vs >4"),
  lty = c(NA, 1, 2, 3, 4), pch = c(1, NA, NA, NA, NA), lwd = 2, cex = 1, bty = "n", col = mycolors[1:5]
)
```



### Fig 9.3: Non-linearities in small sample (n=751); and full GUSTO-I (n=40,830)

```{r Fig9.3, message = FALSE, warning=FALSE }
# Examine non-linearities in gustos sample
# Age
age.linear <- lrm(DAY30 ~ AGE, data = gustos, x = T, y = T, linear.predictors = F)
age.rcs1 <- lrm(DAY30 ~ rcs(AGE, 5), data = gustos, x = T, y = T, linear.predictors = F)
age.fp1 <- mfp(DAY30 ~ fp(AGE, df = 4), alpha = 1, data = gustos, family = binomial) # selected: -2 and 3
age.gam1 <- gam(DAY30 ~ s(AGE), data = gustos, family = binomial)

# examine predictions for age 20:95
age.mat <- matrix(nrow = 751, ncol = 5)
names(age.mat) <- list(NULL, Cs(AGE, linear, fp, rcs, gam))
AGE <- seq(20, 95, by = 0.1)
age.mat[, 1] <- AGE
age.mat[, 2] <- predict(age.linear, newdata = as.data.frame(x = AGE), type = "lp")
age.mat[, 3] <- predict(age.fp1, newdata = as.data.frame(x = AGE))
age.mat[, 4] <- predict(age.rcs1, newdata = as.data.frame(x = AGE), type = "lp")
age.mat[, 5] <- predict(age.gam1, newdata = as.data.frame(x = AGE))

# Plot for n=785, Fig 9.3, part A #
par(mfrow = c(1, 2))
plot(
  x = age.mat[, 1], y = age.mat[, 2], xlim = c(20, 92), ylim = c(-8, 0.2), las = 1, xaxt = "n",
  xlab = "Age in years", ylab = "logit of 30-day mortality", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[2]
)
axis(1, at = c(30, 40, 50, 60, 70, 80, 90))
lines(x = age.mat[, 1], y = age.mat[, 3], lty = 2, lwd = 2, col = mycolors[3])
lines(x = age.mat[, 1], y = age.mat[, 4], lty = 3, lwd = 3, col = mycolors[4])
lines(x = age.mat[, 1], y = age.mat[, 5], lty = 4, lwd = 3, col = mycolors[5])
histSpike(x = gustos$AGE, side = 1, frac = .1, col = "darkblue", add = T)
legend("topleft",
  legend = c("linear", "fp", "rcs", "gam"),
  lty = 1:4, lwd = 2, cex = 1.2, bty = "n", col = mycolors[2:5]
)
title("GUSTO-I, n=785")
# End plot n=785

####################
## Now: N=40,830 ##
age.linear.2 <- lrm(DAY30 ~ AGE, data = gusto, x = T, y = T, linear.predictors = F)
age.rcs1.2 <- lrm(DAY30 ~ rcs(AGE, 5), data = gusto, x = T, y = T, linear.predictors = F)
age.fp1.2 <- mfp(DAY30 ~ fp(AGE, df = 4), alpha = 1, data = gusto, family = binomial) # selected: -2 and 3
age.gam1.2 <- gam(DAY30 ~ s(AGE), data = gusto, family = binomial)

# examine predictions for age 20:95
age.mat.2 <- matrix(nrow = 751, ncol = 5)
names(age.mat.2) <- list(NULL, Cs(AGE, linear, fp, rcs, gam))
age.mat.2[, 1] <- AGE
age.mat.2[, 2] <- predict(age.linear.2, newdata = as.data.frame(x = AGE), type = "lp")
age.mat.2[, 3] <- predict(age.fp1.2, newdata = as.data.frame(x = AGE))
age.mat.2[, 4] <- predict(age.rcs1.2, newdata = as.data.frame(x = AGE), type = "lp")
age.mat.2[, 5] <- predict(age.gam1.2, newdata = as.data.frame(x = AGE))

# Plot for n=40830
plot(
  x = age.mat.2[, 1], y = age.mat.2[, 2], xlim = c(20, 92), ylim = c(-8, 0.2), las = 1, xaxt = "n",
  xlab = "Age in years", ylab = "", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[2]
)
axis(1, at = c(30, 40, 50, 60, 70, 80, 90))
lines(x = age.mat.2[, 1], y = age.mat.2[, 3], lty = 2, lwd = 2, col = mycolors[3])
lines(x = age.mat.2[, 1], y = age.mat.2[, 4], lty = 3, lwd = 3, col = mycolors[4])
lines(x = age.mat.2[, 1], y = age.mat.2[, 5], lty = 4, lwd = 3, col = mycolors[5])
histSpike(x = gusto$AGE, side = 1, frac = .2, col = "darkblue", add = T)
title("GUSTO-I, n=40,830")
```


### Fig 9.4: Glucose and hb in TBI

We evaluate the predictors 'glucose' and 'hemoglobin' in TBI. Both are first truncated, after inspecting boxplots (Fig 9.4), and then analyzed for their prognostic value, with plots for illustration of the estimated relations (Fig 9.5).

```{r Fig9.4, warning=FALSE }
# glucose
quantile(TBI$glucose, probs = c(.005, .01, .02, .98, .99, .995), na.rm = T)
#  0.5%    1%    2%   98%   99% 99.5%
#  1.57  2.26  4.28 18.34 20.92 23.39
# So, winsorize / truncate at 3 and 20
# use simple function to do the winsorizing:
# {ifelse(x<lower,upper,  ifelse(x>upper,upper, x))}
winsorize <- function(x, lower = quantile(x, probs = 0.01), upper = quantile(x, probs = 0.99)) {
  ifelse(x < lower, lower,
    ifelse(x > upper, upper, x)
  )
}
TBI$glucoset <- winsorize(TBI$glucose, 3, 20)

# systolic bp
quantile(TBI$hb, probs = c(.005, .01, .02, .98, .99, .995), na.rm = T)
#  0.5%    1%    2%   98%   99% 99.5%
#  4.87  5.40  6.30 16.50 16.80 17.00
# So, winsorize / truncate at 6 and 17
TBI$hbt <- winsorize(TBI$hb, 6, 17)

# boxplots for illustration
par(mfrow = c(1, 2))
boxplot(x = cbind(TBI$glucose, TBI$glucoset), outcol = c("red", mycolors[4]), border = mycolors[c(10, 4)], xaxt = "n", ylab = "glucose (mmol/L)")
axis(1, at = c(1, 2), labels = c("before ", "after winsorizing"))
title("Glucose")
boxplot(x = cbind(TBI$hb, TBI$hbt), outcol = c("red", mycolors[4]), border = mycolors[c(10, 4)], xaxt = "n", ylab = "hb (mmol/L)")
axis(1, at = c(1, 2), labels = c("before ", "after winsorizing"))
title("Hemoglobin")
```

### Fig 9.5:  Non-linear association of glucose

```{r Fig9.5, message = FALSE, echo=FALSE }
g1 <- lrm(d.gos < 2 ~ rcs(glucose), data = TBI, x = T, y = T)
g2 <- lrm(d.gos < 3 ~ rcs(glucose), data = TBI)
g3 <- lrm(d.gos < 4 ~ rcs(glucose), data = TBI)
g4 <- lrm(d.gos < 5 ~ rcs(glucose), data = TBI)

g1t <- lrm(d.gos < 2 ~ rcs(glucoset), data = TBI)
g2t <- lrm(d.gos < 3 ~ rcs(glucoset), data = TBI)
g3t <- lrm(d.gos < 4 ~ rcs(glucoset), data = TBI)
g4t <- lrm(d.gos < 5 ~ rcs(glucoset), data = TBI)

# data in matrix for plot
glucose <- seq(.5, 41.4, by = 0.1) # 410 elements
g.mat <- matrix(nrow = 410, ncol = 5)
names(g.mat) <- list(NULL, Cs(glucose, g1, g2, g3, g4))
g.mat[, 1] <- glucose
g.mat[, 2] <- predict(g1, newdata = as.data.frame(x = glucose))
g.mat[, 3] <- predict(g2, newdata = as.data.frame(x = glucose))
g.mat[, 4] <- predict(g3, newdata = as.data.frame(x = glucose))
g.mat[, 5] <- predict(g4, newdata = as.data.frame(x = glucose))

# Plots
par(mfrow = c(1, 2))
plot(
  x = g.mat[, 1], y = g.mat[, 2], xlim = c(0, 42), ylim = c(-2, 2), las = 1, xaxt = "n",
  xlab = "Glucose before winsorizing", ylab = "logit GOS at 6 months", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[1]
)
axis(1, at = c(0, 10, 20, 30, 40))
lines(x = g.mat[, 1], y = g.mat[, 3], lty = 2, lwd = 2, col = mycolors[2])
lines(x = g.mat[, 1], y = g.mat[, 4], lty = 3, lwd = 3, col = mycolors[3])
lines(x = g.mat[, 1], y = g.mat[, 5], lty = 4, lwd = 3, col = mycolors[4])
histSpike(x = TBI$glucose, side = 3, frac = .2, col = "darkblue", add = T)
legend("bottomright", legend = c("<good recovery", "unfavorable", "mort+veg", "mortality"), lty = 4:1, lwd = 2, cex = 0.8, bty = "n", col = mycolors[4:1])

# data in matrix for plot 2
glucoset <- seq(3, 20, by = 0.1) # 171 elements
g.mat2 <- matrix(nrow = 171, ncol = 5)
names(g.mat) <- list(NULL, Cs(glucose, g1, g2, g3, g4))
g.mat2[, 1] <- glucoset
g.mat2[, 2] <- predict(g1t, newdata = as.data.frame(x = glucoset))
g.mat2[, 3] <- predict(g2t, newdata = as.data.frame(x = glucoset))
g.mat2[, 4] <- predict(g3t, newdata = as.data.frame(x = glucoset))
g.mat2[, 5] <- predict(g4t, newdata = as.data.frame(x = glucoset))
# Plot
plot(
  x = g.mat2[, 1], y = g.mat2[, 2], xlim = c(0, 20), ylim = c(-2, 2), las = 1, xaxt = "n",
  xlab = "and after winsorizing (mmol/L)", ylab = "", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[1]
)
axis(1, at = c(0, 10, 20))
lines(x = g.mat2[, 1], y = g.mat2[, 3], lty = 2, lwd = 2, col = mycolors[2])
lines(x = g.mat2[, 1], y = g.mat2[, 4], lty = 3, lwd = 3, col = mycolors[3])
lines(x = g.mat2[, 1], y = g.mat2[, 5], lty = 4, lwd = 3, col = mycolors[4])
histSpike(x = TBI$glucoset, side = 3, frac = .2, col = "darkblue", add = T)
```

### Fig 9.6: Systolic blood pressure in TBI

We now examine the prognostic value of systolic blood pressure (BP) in TBI patients. 
We expect low BP (hypotension) to be especially risky.

```{r Fig9.6, message = FALSE}
quantile(TBI$d.sysbpt, probs = c(.01, .25, .75, .99), na.rm = T)
TBIs <- TBI[!is.na(TBI$d.sysbpt), ]
#   1%   25%   75%   99%
# 92.9 121.1 142.1 171.8

g1 <- lrm(d.gos < 2 ~ rcs(d.sysbpt, 3), data = TBIs)
g2 <- lrm(d.gos < 3 ~ rcs(d.sysbpt, 3), data = TBIs)
g3 <- lrm(d.gos < 4 ~ rcs(d.sysbpt, 3), data = TBIs)
g4 <- lrm(d.gos < 5 ~ rcs(d.sysbpt, 3), data = TBIs)

# Define categorical variants of systolic BP
TBIs$d.sysbpt.c <- as.factor(ifelse(TBIs$d.sysbpt < 120, 1, ifelse(TBIs$d.sysbpt > 150, 2, 0)))

g1t <- lrm(d.gos < 2 ~ d.sysbpt.c, data = TBIs)
g2t <- lrm(d.gos < 3 ~ d.sysbpt.c, data = TBIs)
g3t <- lrm(d.gos < 4 ~ d.sysbpt.c, data = TBIs)
g4t <- lrm(d.gos < 5 ~ d.sysbpt.c, data = TBIs)

dd <- datadist(TBIs)
options(datadist = "dd")
# Odds ratios
exp(coef(g1t)) # OR 2.78 for low BP and 1.25 for high BP

describe(TBI$d.sysbpt)
describe(TBIs$d.sysbpt.c)
# data in matrix for plot
d.sysbpt <- seq(60, 210, by = 1) # 151 elements
g.mat <- matrix(nrow = 151, ncol = 5)
names(g.mat) <- list(NULL, Cs(d.sysbpt, g1, g2, g3, g4))
g.mat[, 1] <- d.sysbpt
g.mat[, 2] <- predict(g1, newdata = as.data.frame(x = d.sysbpt))
g.mat[, 3] <- predict(g2, newdata = as.data.frame(x = d.sysbpt))
g.mat[, 4] <- predict(g3, newdata = as.data.frame(x = d.sysbpt))
g.mat[, 5] <- predict(g4, newdata = as.data.frame(x = d.sysbpt))

# Plot: Fig 9.6, part I
par(mfrow = c(1, 2))
plot(
  x = g.mat[, 1], y = g.mat[, 2], xlim = c(50, 210), ylim = c(-2.7, 2.7), las = 1, xaxt = "n",
  xlab = "Systolic blood pressure (mmHg)", ylab = "logit GOS at 6 months", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[1]
)
axis(1, at = c(100, 150, 200))
lines(x = g.mat[, 1], y = g.mat[, 3], lty = 2, lwd = 2, col = mycolors[2])
lines(x = g.mat[, 1], y = g.mat[, 4], lty = 3, lwd = 3, col = mycolors[3])
lines(x = g.mat[, 1], y = g.mat[, 5], lty = 4, lwd = 3, col = mycolors[4])
histSpike(x = TBI$d.sysbpt, side = 3, frac = .2, col = "darkblue", add = T)
legend("bottomleft", legend = c("<good recovery", "unfavorable", "mort+veg", "mortality"), lty = 4:1, lwd = 2, cex = 0.8, bty = "n", col = mycolors[4:1])
title("continuous, rcs 3 knots")

## Plot 9.6, part II
d.sysbpt.c <- c(0, 1, 2)
d.sysbpt <- seq(60, 210, by = .1) # 1501 elements
g.mat2 <- matrix(nrow = 1501, ncol = 5)
g1t.c <- predict(g1t, newdata = as.data.frame(x = d.sysbpt.c))
g2t.c <- predict(g2t, newdata = as.data.frame(x = d.sysbpt.c))
g3t.c <- predict(g3t, newdata = as.data.frame(x = d.sysbpt.c))
g4t.c <- predict(g4t, newdata = as.data.frame(x = d.sysbpt.c))

names(g.mat2) <- list(NULL, Cs(d.sysbpt, g1, g2, g3, g4))
g.mat2[, 1] <- d.sysbpt
g.mat2[, 2] <- ifelse(d.sysbpt < 120, g1t.c[2], ifelse(d.sysbpt > 150, g1t.c[3], g1t.c[1]))
g.mat2[, 3] <- ifelse(d.sysbpt < 120, g2t.c[2], ifelse(d.sysbpt > 150, g2t.c[3], g2t.c[1]))
g.mat2[, 4] <- ifelse(d.sysbpt < 120, g3t.c[2], ifelse(d.sysbpt > 150, g3t.c[3], g3t.c[1]))
g.mat2[, 5] <- ifelse(d.sysbpt < 120, g4t.c[2], ifelse(d.sysbpt > 150, g4t.c[3], g4t.c[1]))

# Plot
plot(
  x = g.mat2[, 1], y = g.mat2[, 2], xlim = c(50, 210), ylim = c(-2.7, 2.7), las = 1, xaxt = "n",
  xlab = "Systolic blood pressure (mmHg)", ylab = "", cex.lab = 1.2, type = "l", lwd = 2, col = mycolors[1]
)
axis(1, at = c(100, 150, 200))
lines(x = g.mat2[, 1], y = g.mat2[, 3], lty = 2, lwd = 2, col = mycolors[2])
lines(x = g.mat2[, 1], y = g.mat2[, 4], lty = 3, lwd = 3, col = mycolors[3])
lines(x = g.mat2[, 1], y = g.mat2[, 5], lty = 4, lwd = 3, col = mycolors[4])
histSpike(x = TBI$d.sysbpt, side = 3, frac = .2, col = "darkblue", add = T)
title("3 categories")
```
