# Assumptions in Regression Models - Additivity and Linearity {#additivity-linearity}

## GUSTO-I interaction analysis

```{r setup-ch12, include=FALSE}
knitr::opts_knit$set(
  echo = TRUE,
  root.dir = here::here()
)

knitr::opts_chunk$set(
  fig.path = "fig/"
)

# Load chapter libraries
library(foreign)
library(rms)
library(interactions)
library(mfp) # the MFP library
library(gam)
library(mgcv)
library(kableExtra)

# Colours
mycolors <- c(
  "black" = 1,
  "Red" = "#ED0000",
  "CongressBlue" = "#00468B",
  "Apple" = "#42B540",
  "BondiBlue" = "#0099B4",
  "TrendyPink" = "#925E9F",
  "Carmine" = "#AD002A",
  "CodGray" = "#1B1919",
  "MonaLisa" = "#FDAF91",
  "Edward" = "#ADB6B6"
)
```

### Examine interactions


```{r}
# Import gusto, gustoB, and sample4 data sets
gusto <- read.csv("data/gusto1.csv") # GUSTO sample with 40830 patients
gustoB <- read.csv("data/gustoB.csv") # GUSTO part B sample with 20318 patients
gustos <- read.csv("data/sample4.csv") # GUSTO sample4 with 785 patients

source("R/auc.nonpara.mw.R")
source("R/ci.auc.R")
source("R/val.prob.ci.2.R")

# levels(gustos$HRT) <- c("No tachycardia", "Tachycardia")
dd <- datadist(gustos)
options(datadist = "dd")
```


Evaluate interactions with age in a full model, which includes 8 predictors in total. The data set is small (sample4, n=785, 52 events)

```{r interactions}
### Full model and age interactions
full <- lrm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + HRT + TTR + SEX, data = gustos, x = T, y = T, linear.predictors = F)
fullint <- lrm(DAY30 ~ AGE * (KILLIP + HIG + DIA + HYP + HRT + TTR + SEX), data = gustos, x = T, y = T, linear.predictors = F)
anova(fullint)

### Select only interaction AGE * HRT
fullints <- lrm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + HRT + TTR + SEX + AGE * HRT, data = gustos, x = T, y = T, linear.predictors = T)
anova(fullints)
```


### Fig 12.1

Make 2 plots with linear interaction, in the small n=785 sample, and in the full n=40830 sample

``` {r Fig 12.1, echo=F}
### Linear interaction
fullints2 <- glm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + HRT + TTR + SEX + AGE * HRT, data = gustos, family = "binomial")
fullintF <- update(fullints2, data = gusto)

### Plot in n=785
interact_plot(fullints2,
  pred = AGE, modx = HRT, outcome.scale = "link", colors = mycolors[c(4, 2)], lwd = c(3, 2), main.title = "GUSTO-I, n=785, AGE*HRT interaction",
  interval = T, xlim = c(35, 85), x.label = "Age (years)", y.label = "Logit 30-day mortality", modx.labels = c(0, 1), legend.main = "HRT"
)
### Plot in full data, n=40830
interact_plot(fullintF,
  pred = AGE, modx = HRT, outcome.scale = "link", colors = mycolors[c(4, 2)], lwd = c(3, 2), main.title = "GUSTO-I, n=40,830, AGE*HRT interaction",
  interval = T, ylim = c(-4, 0), xlim = c(35, 85), x.label = "Age (years)", y.label = "Logit 30-day mortality", modx.labels = c(0, 1), legend.main = "HRT"
)
## End Fig 12.1
```


### Fig 12.2

Make 4 plots with main effects, linear interaction, and 2 variants of interaction only above age 55. The variable is (Age-55)[+].

``` {r Fig 12.2, echo=F, warning=F}
### Main effects
full8 <- glm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + HRT + TTR + SEX, data = gustos, family = "binomial")
interact_plot(full8,
  pred = AGE, modx = HRT, outcome.scale = "link", colors = mycolors[c(4, 2)], lwd = c(3, 2), main.title = "AGE + HRT, 2 df",
  interval = T, xlim = c(35, 85), x.label = "Age (years)", y.label = "Logit 30-day mortality", modx.labels = c(0, 1), legend.main = "HRT"
)

### Linear interaction
fullints2 <- update(full8, . ~ . + AGE * HRT)
interact_plot(fullints2,
  pred = AGE, modx = HRT, outcome.scale = "link", colors = mycolors[c(4, 2)], lwd = c(3, 2), main.title = "GUSTO-I, n=785, AGE*HRT interaction",
  interval = T, xlim = c(35, 85), x.label = "Age (years)", y.label = "Logit 30-day mortality", modx.labels = c(0, 1), legend.main = "HRT"
)
### Age>55 interaction, 3 df
fullints4 <- glm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + TTR + SEX + ifelse(AGE > 55, (AGE - 55) * (HRT), 0) + ifelse(AGE > 55, (AGE - 55) * (1 - HRT), 0), data = gustos, family = "binomial")
interact_plot(fullints4,
  pred = AGE, modx = HRT, data = gustos, outcome.scale = "link", colors = mycolors[c(4, 2)], lwd = c(3, 2),
  main.title = ("AGE + (AGE-55))[+] * HRT + (AGE-55)[+] * (1-HRT), 3 df"),
  interval = T, ylim = c(-9, 1), xlim = c(35, 85), x.label = "Age (years)", y.label = "Logit 30-day mortality", modx.labels = c(0, 1), legend.main = "HRT"
)

### Age>55 interaction, 2 df; no effect of age among HRT=0
fullints3 <- glm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + TTR + SEX + ifelse(AGE > 55, (AGE - 55) * (HRT), 0), data = gustos, family = "binomial")
interact_plot(fullints3,
  pred = AGE, modx = HRT, data = gustos, outcome.scale = "link", colors = mycolors[c(4, 2)], lwd = c(3, 2),
  main.title = "AGE + (AGE-55)[+] * HRT, 2 df",
  interval = T, x.label = "Age (years)", y.label = "Logit 30-day mortality", modx.labels = c(0, 1), legend.main = "HRT"
)

## End Fig 12.2
```

In the last graph, the green dotted line follows the angle from below age 55 years (only 1 Age effect is estimated for the HRT==0 and HRT==1 and age<55 patients). In the pre-final graph, there are 2 separate angles from age 55 for HRT==0 and HRT==1 (barely noticable for the green dotted line).


### Smart coding illustration

Smart coding of age effect: separate for no HRT (HRT==0) and for HRT (HRT==1)

``` {r Coding, echo=T, warning=F}
# Smart coding of age effect: separate for no HRT (HRT==0) and for HRT (HRT==1)
gustos$AGE0 <- gustos$AGE * (1 - gustos$HRT)
gustos$AGE1 <- gustos$AGE * gustos$HRT
# Standard
lrm(DAY30 ~ AGE + KILLIP + HIG + DIA + HYP + HRT + TTR + SEX + AGE * HRT, data = gustos, x = T, y = T, linear.predictors = F)
# Smart coding
lrm(DAY30 ~ AGE0 + AGE1 + HRT + KILLIP + HIG + DIA + HYP + TTR + SEX, data = gustos, x = T, y = T, linear.predictors = F)
# Identical fit, easier interpretation

# Age 55 as reference for HRT effect
gustos$AGE0 <- (gustos$AGE - 55) * (1 - gustos$HRT)
gustos$AGE1 <- (gustos$AGE - 55) * gustos$HRT
lrm(DAY30 ~ AGE0 + AGE1 + HRT + KILLIP + HIG + DIA + HYP + TTR + SEX, data = gustos, x = T, y = T, linear.predictors = F)
# Even nicer interpretation, HRT effect for age=55
```

The fit of each of the models is identical (always LR chi2=86.28); each model allows for linear interaction between AGE and HRT. The interpretation of the model is easier if the age effects are estimated for HRT==1 and for HRT==0. Scaling is easier by subtracting 55 from AGE (AGE-55); this implies the HRT effect relates to age 55.


### Table 12.2 Better predictions? 

Assess the performance if the models created in n=785 in an independent validation part, GustoB, n=20318.  
Plots created with a modification of Frank Harrell's val.prob() function:  

```{r, results='hide',fig.keep='all'}
# Validate in independent part, named gustoB
# main effects
lrm.val.full <- predict(full, newdata = gustoB, type = "lp")
# simple interaction
lrm.val.int1 <- predict(fullints, newdata = gustoB, type = "lp")

# Plot
val.prob.ci.2(
  y = gustoB[, "DAY30"], logit = lrm.val.full, riskdist = "predicted", logistic.cal = F,
  smooth = "rcs", nr.knots = 3, g = 8, xlim = c(0, .5), ylim = c(0, .5),
  legendloc = c(0.18, 0.15), statloc = c(0, .4), roundstats = 3,
  xlab = "Predicted probability from n=785", ylab = "Observed proportion in n=20318"
)
```   

``` {r Table 12.2, echo=F, results=F, warning=F}
# Age>55 interaction, 3 df, or 2 df
gustos$AGE55min <- ifelse(gustos$AGE < 55, gustos$AGE - 55, 0)
gustos$AGE55plusHRT <- ifelse(gustos$AGE < 55 | gustos$HRT == 0, 0, gustos$AGE - 55)
gustos$AGE55plusNoHRT <- ifelse(gustos$AGE < 55 | gustos$HRT == 1, 0, gustos$AGE - 55)
gustos$AGE55 <- gustos$AGE - 55
# Same for GustoB set
gustoB$AGE55min <- ifelse(gustoB$AGE < 55, gustoB$AGE - 55, 0)
gustoB$AGE55plusHRT <- ifelse(gustoB$AGE < 55 | gustoB$HRT == 0, 0, gustoB$AGE - 55)
gustoB$AGE55plusNoHRT <- ifelse(gustoB$AGE < 55 | gustoB$HRT == 1, 0, gustoB$AGE - 55)
gustoB$AGE55 <- gustoB$AGE - 55

# Fit in n=785
fit3ages <- lrm(DAY30 ~ AGE55min + AGE55plusNoHRT + AGE55plusHRT + KILLIP + HIG + DIA + HYP + TTR + SEX, data = gustos, x = T, y = T, linear.predictors = F)
fitage.HRT <- lrm(DAY30 ~ AGE55 + AGE55plusHRT + KILLIP + HIG + DIA + HYP + TTR + SEX, data = gustos, x = T, y = T, linear.predictors = F)

# Age>55 interaction, 3 df
lrm.val.int2 <- predict(fit3ages, newdata = gustoB, type = "lp")
# Age>55 interaction, 2 df; no effect of age among HRT=0
lrm.val.int3 <- predict(fitage.HRT, newdata = gustoB, type = "lp")

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
val.prob.ci.2(
  y = gustoB[, "DAY30"], logit = lrm.val.full, riskdist = "predicted", logistic.cal = F, smooth = "rcs", nr.knots = 3, g = 8,
  xlim = c(0, .5), ylim = c(0, .5), legendloc = c(0.18, 0.15), statloc = c(0, .4), roundstats = 3,
  xlab = "Predicted probability from n=785", ylab = "Observed proportion in n=20318"
)
title(bquote(paste("Main effects, 2", italic("df"))))

val.prob.ci.2(
  y = gustoB[, "DAY30"], logit = lrm.val.int1, riskdist = "predicted", logistic.cal = F, smooth = "rcs", nr.knots = 3, g = 8,
  xlim = c(0, .5), ylim = c(0, .5), legendloc = c(0.18, 0.15), statloc = c(0, .4), roundstats = 3,
  xlab = "Predicted probability from n=785", ylab = "Observed proportion in n=20318"
)
title(bquote(paste("Simple AGE*HRT interaction, 3", italic("df"))))

val.prob.ci.2(
  y = gustoB[, "DAY30"], logit = lrm.val.int2, riskdist = "predicted", logistic.cal = F, smooth = "rcs", nr.knots = 3, g = 8,
  xlim = c(0, .5), ylim = c(0, .5), legendloc = c(0.18, 0.15), statloc = c(0, .4), roundstats = 3,
  xlab = "Predicted probability from n=785", ylab = "Observed proportion in n=20318"
)
title(bquote(paste("(AGE-55)+*HRT interaction, 3", italic("df"))))

val.prob.ci.2(
  y = gustoB[, "DAY30"], logit = lrm.val.int3, riskdist = "predicted", logistic.cal = F, smooth = "rcs", nr.knots = 3, g = 8,
  xlim = c(0, .5), ylim = c(0, .5), legendloc = c(0.18, 0.15), statloc = c(0, .4), roundstats = 3,
  xlab = "Predicted probability from n=785", ylab = "Observed proportion in n=20318"
)
title(bquote(paste("AGE + (AGE-55)+*HRT, 2", italic("df"))))
# End performance evaluation in GustoB part, n=20318
```

**Conclusions**
Discrimination: worse with interactions than without.
Calibration: We note some overfitting, as expected by a fit in a small sample.

## MFP and other non-linear analyses in n544 data

Upcoming.
